---
title: 'PSTAT 220A HW #1'
author: "Jaxon Stuhr"
date: "2022-09-27"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(ggplot2)
library(faraway)
library(car)
library(MASS)
```

# Question 1
```{r, echo=TRUE}
# this function returns the standard deviation, mean absolute deviation, median absolute deviation, and inter-quartile range of dataset_x
dispersion_measures = function(dataset_x) {
  sd = sd(dataset_x) # standard deviation
  mean_ad = mad(dataset_x, center = mean(dataset_x)) # mean absolute deviation
  median_ad = mad(dataset_x, center = median(dataset_x)) # median absolute deviation
  iqr = IQR(dataset_x) # inter-quartile range
  return(c(sd, mean_ad, median_ad, iqr))
}
```

# Question 2
```{r}
pima_dataset = pima

# dispersion measures for each paramters
vals_pregnant = dispersion_measures(pima_dataset$pregnant)
vals_diastolic = dispersion_measures(pima_dataset$diastolic)
vals_bmi = dispersion_measures(pima_dataset$bmi)
vals_age = dispersion_measures(pima_dataset$age)

# "pregnant" plots
par(mfrow=c(2,2))
boxplot(pima_dataset$pregnant, main = "Boxplot", ylab = "Number of Times Pregnant")
hist(pima_dataset$pregnant, main = "Histogram", xlab = "Number of Times Pregnant", ylab = "Count")
qqPlot(pima_dataset$pregnant, main = "QQ-Plot", ylab = "Sample Quantiles", xlab = "Theoretical Quantiles")
d = density(pima_dataset$pregnant)
plot(d, 
     xlim = c(-5,20), 
     main = "Kernel Density Estimate", 
     xlab="Times Pregnant",
     ylab = "Density")

# "diastolic" plots
par(mfrow=c(2,2))
boxplot(pima_dataset$diastolic, main = "Boxplot", ylab = "Diastolic blood pressure (mm Hg)")
hist(pima_dataset$diastolic, main = "Histogram", xlab = "Diastolic blood pressure (mm Hg)", ylab = "Count")
qqPlot(pima_dataset$diastolic, main = "QQ-Plot", ylab = "Sample Quantiles", xlab = "Theoretical Quantiles")
d = density(pima_dataset$diastolic)
plot(d, 
     xlim = c(-10,10), 
     main = "Kernel Density Estimate", 
     xlab="Diastolic blood pressure (mm Hg)",
     ylab = "Density")

# "bmi" plots
par(mfrow=c(2,2))
boxplot(pima_dataset$bmi, main = "Boxplot", ylab = "Body Mass Index (BMI)")
hist(pima_dataset$bmi, main = "Histogram", xlab = "Body Mass Index (BMI)", ylab = "Count")
qqPlot(pima_dataset$bmi, main = "QQ-Plot", ylab = "Sample Quantiles", xlab = "Theoretical Quantiles")
d = density(pima_dataset$bmi)
plot(d, 
     xlim = c(0,90), 
     main = "Kernel Density Estimate", 
     xlab="Body Mass Index (BMI)",
     ylab = "Density")

# "age" plots
par(mfrow=c(2,2))
boxplot(pima_dataset$age, main = "Boxplot", ylab = "Age (years)")
hist(pima_dataset$age, main = "Histogram", xlab = "Age (years)", ylab = "Count")
qqPlot(pima_dataset$age, main = "QQ-Plot", ylab = "Sample Quantiles", xlab = "Theoretical Quantiles")
d = density(pima_dataset$age)
plot(d, 
     xlim = c(0,90), 
     main = "Kernel Density Estimate", 
     xlab="Age (years)",
     ylab = "Density")

```

Pregnant: a few very high outliers, distribution appears to be Gamma

Blood Pressure: distribution looks normal except for outliers with extremely low bloodpressure

BMI: distribution also looks relatively normal, but again there are extreme outliers with very low BMI that fall outside the distribution

Age: looks like a gamma distribution, or exponential after year 18... Definitely not normal due to QQ-plot being highly non-linear


# Question 3
```{r}
precip = data.frame(vals = (c(0.77, 1.20, 3.00, 1.62, 2.81, 2.48, 1.74, 0.47, 
                      3.09, 1.31, 1.87, 0.96, 0.81, 1.43, 1.51, 0.32, 
                      1.18, 1.89, 1.20, 3.37,2.10, 0.59, 1.35, 0.90, 
                      1.95, 2.20, 0.52, 0.81, 4.75, 2.05)) )
qqPlot(precip$vals, xlab = "Theoretical Quantiles", ylab = "Sample Quantiles")
```

Looks farily normal, but tails are not linear. Try log(precip)...

```{r}
precip = precip %>% 
  mutate(log_vals = log(vals))

qqPlot(precip$log_vals, xlab = "Theoretical Quantiles", ylab = "log(Sample) Quantiles")
```

That looks like a normal distribution!

# Question 4

```{r}
set.seed(127)

failure_times = data.frame(times = c(4381, 3953, 2603, 2320, 1161, 3286, 6914,
                                     4007, 3168, 2376, 7498, 3923 ,9460 ,4525,
                                     2168, 1288, 5085, 2217, 6922, 218, 1309,
                                     1875, 1023, 1697, 1038, 3699, 6142,4732,
                                     3330, 4159 ,2537, 3814, 2157, 
                                     7683, 5539,4839,6052,
                                     2420, 5556, 309, 1295, 3266, 6679, 
                                     1711, 5931))

points = ppoints(failure_times$times)

par(mfrow=c(2,2))

# Normal Distribution fit and QQ-plot
fit.normal = fitdistr(failure_times$times/100, 
                      "normal", 
                      lower = 0.001)
plot(qnorm(points, mean = fit.normal$estimate[1], sd = fit.normal$estimate[2]),
     sort(failure_times$times/100),
     xlim = c(0,100), 
     ylim = c(0,100),
     main = "Normal Q-Q plot",
     xlab="Theoretical Quantiles",
     ylab = "Sample Quantiles"
     )
abline(0,1)

# Lognormal Distribution fit and QQ-plot
fit.lognormal = fitdistr(failure_times$times/100, 
                      "lognormal", 
                      lower = 0.001)
plot(qlnorm(points, meanlog = fit.lognormal$estimate[1], sdlog = fit.lognormal$estimate[2]),
     sort(failure_times$times/100),
     xlim = c(0,120), 
     ylim = c(0,120),
     main = "Lognormal Q-Q plot",
     xlab="Theoretical Quantiles",
     ylab = "Sample Quantiles"
     )
abline(0,1)

# Exponential Distribution fit and QQ-plot
fit.exponential = fitdistr(failure_times$times/100, 
                      "exponential", 
                      lower = 0.001)
plot(qexp(points, rate = fit.exponential$estimate[1]),
     sort(failure_times$times/100),
     xlim = c(0,120), 
     ylim = c(0,120),
     main = "Exponential Q-Q plot",
     xlab="Theoretical Quantiles",
     ylab = "Sample Quantiles"
     )
abline(0,1)

# Gamma Distribution fit and QQ-plot
fit.gamma <- fitdistr(failure_times$times/100, 
                      "gamma", 
                      list(shape=1,rate=1),
                      lower = 0.001)
plot(qgamma(points, shape = fit.gamma$estimate[1], rate = fit.gamma$estimate[2]), 
     sort(failure_times$times/100),
     xlim = c(0,120), 
     ylim = c(0,120),
     main = "Gamma Q-Q plot",
     xlab="Theoretical Quantiles",
     ylab = "Sample Quantiles")
abline(0,1)

```

Based on the above plots, the Gamma distribution appears to fit best as the points closely follow the line, except for the last two outliers. The Normal distribution fits relatively well, and the exponential and lognormal are very poor fits.

```{r}
set.seed(127)
# compare Gamma PDF with Kernel Density Estimate
par(mfrow=c(1,2))
# plot Kernel density estimate
d = density(failure_times$times/100)
plot(d, 
     xlim = c(0,100), 
     main = "Kernel Density Estimate", 
     xlab="Failure Times",
     ylab = "Density")
# plot gamma density function
xgamma = seq(0,100,by = 1)
plot(dgamma(xgamma, 
            shape = fit.gamma$estimate[1], 
            rate = fit.gamma$estimate[2]), 
     xlim = c(0,100),
     main = "Gamma Distribution",
     xlab = "Failure Times",
     ylab = "Density")
```

Gamma distribution with given paramters aligns quite well with KDE!

# Question 5

```{r}
# generate exponential samples
exp_samples = rexp(n = 5000, rate = 1)

# plot histogram
hist(exp_samples, main = "Histogram of Samples", xlab = "Sample Value", ylab = "Count")
```

The shape drops of as x increases but has a long tail, as we would expect because the exponential function goes as $f(x) = e^{-x}$

```{r}
# generate samples and means blank matrices
samples = matrix(0, 100, 50)
means = matrix(0, 1, 50)
# build samples matrix from exp_samples
for (x in 1:50) {
  samples[,x] = exp_samples[(1+100*(x-1)):(100*x)]
}
# calculate means for each grouping
for (x in 1:50) {
  sample_mean = mean(samples[, x])
  means[1,x] = sample_mean
}
# plot histogram
hist(means, main = "Histogram of Means", xlab = "Mean Value", ylab = "Frequency")
```

They don't have the same shape, because the larger values that make up the tail of the exponential histogram will be averaged with the higher density of lower values, leaving a more normal distribution around 1. 

# Question 6

```{r}
# generate exponential samples
normal_samples = rnorm(n = 600, mean = 10, sd = 5)
# generate samples and means blank matrices
normal_groups = matrix(0, 6, 100)
stats = matrix(0, 1, 100)
# build samples matrix from exp_samples
for (x in 1:100) {
  normal_groups[,x] = normal_samples[(1+6*(x-1)):(6*x)]
}
# calculate means for each grouping
for (x in 1:100) {
  group_mean = mean(normal_groups[, x])
  group_var = var(normal_groups[,x])
  stats[1,x] = (group_mean - 10)/sqrt(group_var/6)
}
# plot histogram
hist(stats, main = "Histogram of Statistics", xlab = "Statistic Value", ylab = "Frequency")
```

We would expect this statistic to be normally distributed about zero, which it is. 